Inverse conformational selection in lipid-protein binding

### 2021-05-29 CoverLetter.tex
Update manuscript and cover letter

Please find enclosed our manuscript entitled {\it Conformational plasticity of lipid headgroups in cellular membranes and protein-lipid complexes}
by A. Bacle et al. which we would like to submit for consideration to be published in \textit{ACS Central Science}.

In the manuscript, we demonstrate that lipids in cellular membranes exhibit a much higher plasticity in terms of allowed conformational states than previously anticipated.
Our results update the current understanding on how lipids selectively bind to proteins.

Lipids with different headgroups, the water facing components of cell membranes, are known to regulate cell functions
by selective binding to different proteins. It is, however, not known if the selectivity is driven by unique
lipid conformations between the lipid types that fit to binding pockets in proteins (lock and key model), or by some other interactions.
One of the main obstacles in understanding the selective binding has been the lack of
experimental methods to resolve the conformational ensembles of lipids when they are not bound to proteins.

In 2013 we set out to solve the conformational ensembles of lipid headgroups by establishing the NMRlipids open collaboration project
which combines data from molecular dynamics (MD) simulations and solid-state NMR experiments.
Using the extensive data collected within this project, we have now resolved the differences in conformational ensembles
between the most common lipids in eukaryotes and bacteria. 
Our results reveal a significant structural plasticity in lipid headgroups and only minor differences of the conformational ensembles
between lipids with different headgroup chemistry.
Notably, and motivated by our findings on the model systems, by using a bioinformatical analysis of protein-bound lipid structures from the protein data bank (PDB), we found that the plasticity in lipid headgroups is a characteristic feature in lipid-protein complexes.

The novel conclusions from our work are that lipid headgroups can adapt to multiple binding sites in proteins,
and the selectivity of lipid-protein binding is driven by intermolecular interactions rather than by restrictions in lipid conformations.
Furthermore, we demonstrated the power of a conceptually novel open collaboration approach that extended the realm of Molecular Dynamics simulations to complement the experimental data in the PDB for elucidating how complex systems made up of disordered biomolecules behave.

We believe that our approach and findings are of great interest to understand the connection of lipid chemistry to protein-lipid interactions  and ultimately the role of lipid chemistry in a number of diseases. Therefore, we believe that \textit{ACS Central Science} is a highly suitable journal to publish this work.

>>>>

Please find enclosed our manuscript entitled
{\it Inverse conformational selection in lipid-protein binding}
by A. Bacle et al. which we would like to submit for consideration to be published in the \textit{Journal of American Chemical Society}.

In the manuscript, we present the {\it Inverse conformational selection model} for lipid-protein binding.
The model illustrates how lipids in cellular membranes or in lipid nanoparticles can select
suitable conformations from wide ensemble %sampled in physiologically relevant liquid state
to bind in various biomolecules, such as proteins, drugs, RNA or viruses.
Our model is formulated based on solid state NMR experiments, molecular dynamics (MD) simulations,
and bioinformatical analysis.
First, the extensive NMR and MD simulation data collected within the NMRlipids project
(open collaboration project started by us in 2013
with 210 citations in total) revealed that 
the most common lipid types in eukaryotes and bacteria exhibit a much higher plasticity in
terms of allowed conformational states than previously anticipated.
Motivated by these results, 
we used a novel bioinformatical analysis of protein-bound lipid structures from the protein data bank (PDB),
and found that the plasticity in lipid headgroups is a characteristic feature also in lipid-protein complexes.

While lipid-protein interactions, and cell functions regulated by them,  remain
as a central topic in molecular cell biology, interactions of lipids with proteins and other biomolecules
is gaining interested also in nanobiotechnology. For example, lipid nanoparticles used as carriers
in mRNA vaccines against COVID-19 can cause allergic reactions.
Our {\it Inverse conformational selection model} can be used to understand how
these nanoparticles interact with RNA cargo and physiological environment,
thereby facilitating the design of novel vaccine or drug carrier with less side effects.
Furthermore, the {\it Inverse conformational selection model}
may apply also to other than lipid molecules, such as intrinsically disordered proteins or sugars.

We believe that the proposed conceptually novel {\it Inverse conformational selection model} 
contributes to wide scope of research fields, ranging from biophysics and biochemistry to biology and biotechnology.
Moreover, the combination of the state-of-the-art NMR experiments and MD simulations with original open collaboration
and bioinformatical approaches further increase the significance of the manuscript.
%the open access policy of {\it Nature Communications} is optimal for the publications from
%the NMRlipids Project which fosters the open collaboration and open data approaches as all the data and other material
%are publicly available throughout the project.
Therefore, we believe that our manuscript fulfills the high standards for publications in the \textit{Journal of American Chemical Society}.

### 2021-05-29 manuscriptPGPE.tex
Update manuscript and cover letter

\begin{abstract}
% insert abstract here
  Interest in lipid interactions with proteins and other biomolecules is emerging not only in
  fundamental biochemistry, but also in the field of nanobiotechnology where lipid nanoparticles
  are commonly used, for example, as carriers in mRNA vaccines.
  The outward facing components of lipid bilayer, the lipid headgroups, regulate how cellular membranes or lipid nanoparticles
  interact with approaching substances, such as proteins, drugs, RNA or viruses.
  %In the physiologically relevant liquid state,
  %lipid molecules are not fixed to a single structure but sample an ensemble of conformations with Boltzmann-weighted probabilities.
  Because conformational ensembles of lipids in physiologically relevant liquid state 
  %However, the distributions of conformations in such ensembles and their lipid-type dependence
  have not yet been experimentally determined, a crucial piece has been missing from the conception on how they interact with other biomolecules: 
  It is not clear if lipid molecules exchange between a few rigid structures,
  or fluctuate freely across a practically continuous spectrum of  conformations.
  Here, we combine solid-state NMR experiments and molecular dynamics simulations from the NMRlipids Project to resolve the conformational ensembles of headgroups of four key lipid types in various biologically relevant conditions. We find that all lipid headgroup types sample a wide range of conformations % a wide range of conformations
  in both neutral and charged cellular membranes, and that differences in the headgroup chemistry manifests only
  in probability distributions of conformations. %, but all four lipid types can access .
  Together with the analysis of 894 protein-bound lipid structures from the Protein Data Bank (PDB), this suggests that lipids can bind to proteins in a wide range of conformations, which are not limited by the headgroup chemistry.
  %Therefore, the selective adsorption of peripheral proteins to membranes and the binding of lipids to transmembrane proteins are likely to be regulated by specific protein--lipid interactions, not by any inherent conformational restrictions of the lipids.
  %Our results pave the way to comprehensive understanding of lipid--protein interactions in biomedical applications, in particular in
  %lipid-mediated signaling.
  Our results propose that lipids are able to select a suitable conformation from the wide range of selection to fit the various binding sites in proteins.
  The proposed {\it inverse conformational selection model} can be also extended to lipid binding to other biomolecules than proteins, such as drug, RNA and viruses.

@@

Our results suggest that lipid binding to proteins can be described by the {\it inverse conformational selection model}
where unbound lipids sample a wide conformational ensemble offering the large selection of conformations
that can bind to various binding sites in substrates, such as protein, drug, RNA, or virus, that remain conformationally fixed.
In this model, the binding energy is dominated by
intermolecular interactions and changes in lipid conformational entropy, while 
conformational restrictions inherent to the lipids play a minor role.
%We suggest that the key to understanding selective binding of certain lipid types to proteins is in intermolecular lipid--protein interactions, rather than in conformational restrictions inherent to the lipids. The broad conformational ensembles we see in bulk bilayers suggest that the structures observed in pure lipid crystals occur seldomly in liquid state. Consequently, in protein binding the entropic cost on the lipid may be significant.
Applications of the proposed model in molecular cell biology include the understanding of lipid-mediated cell signaling
and functions of lipid regulated membrane proteins in general. 
In bionanotechnology, our model facilitates, for example,
the design of phoshopolipid-specific antibodies and lipid nanoparticle carries with minimal sides effects for mRNA vaccines.
Furthermore, our work has demonstrated the power of open access MD simulation data
to complement the PDB data in elucidating behavior of complex systems of disordered biomolecules.

>>>>

\begin{abstract}
% insert abstract here
  Interest in lipid interactions with proteins and other biomolecules is emerging not only in
  fundamental biochemistry, but also in the field of nanobiotechnology where lipid nanoparticles
  are commonly used, for example, as carriers in mRNA vaccines.
  The outward facing components of lipid bilayer, the lipid headgroups, regulate how cellular membranes or lipid nanoparticles
  interact with approaching substances, such as proteins, drugs, RNA or viruses.
  %In the physiologically relevant liquid state,
  %lipid molecules are not fixed to a single structure but sample an ensemble of conformations with Boltzmann-weighted probabilities.
  Because conformational ensembles of lipids in physiologically relevant liquid state 
  %However, the distributions of conformations in such ensembles and their lipid-type dependence
  have not yet been experimentally determined, a crucial piece has been missing from the understanding
  on how they interact with other biomolecules: 
  It is not clear if lipid molecules exchange between a few rigid structures,
  or fluctuate freely across a practically continuous spectrum of  conformations.
  Here, we combine solid-state NMR experiments and molecular dynamics simulations from the NMRlipids Project to resolve the conformational ensembles of headgroups of four key lipid types in various biologically relevant conditions. We find that all lipid headgroup types sample a wide range of
  overlapping conformations % a wide range of conformations
  in both neutral and charged cellular membranes, and that differences in the headgroup chemistry manifests only
  in probability distributions of conformations. %, but all four lipid types can access .
  Together with the analysis of 894 protein-bound lipid structures from the Protein Data Bank (PDB),
  this suggests that all four key lipid types can bind to proteins in a wide range of conformations,
  which are not limited by the headgroup chemistry.
  %Therefore, the selective adsorption of peripheral proteins to membranes and the binding of lipids to transmembrane proteins are likely to be regulated by specific protein--lipid interactions, not by any inherent conformational restrictions of the lipids.
  %Our results pave the way to comprehensive understanding of lipid--protein interactions in biomedical applications, in particular in
  %lipid-mediated signaling.
  Our results propose that lipids are able to select a suitable conformation
  from the wide range available to them to fit the various binding sites in proteins.
  The proposed {\it Inverse Conformational Selection Model} will extend also to lipid binding
  to other targets than proteins, such as drugs, RNA and viruses.

@@

Our results suggest that lipid binding to proteins can be described by the {\it inverse conformational selection model}
where unbound lipids sample a wide conformational ensemble offering the large selection of conformations
that can bind to various binding sites in substrates, such as protein, drug, RNA, or virus, that remain conformationally fixed.
In this model, the binding energy is dominated by
intermolecular interactions and changes in lipid conformational entropy, while 
conformational restrictions inherent to the lipids play a minor role.
%We suggest that the key to understanding selective binding of certain lipid types to proteins is in intermolecular lipid--protein interactions, rather than in conformational restrictions inherent to the lipids. The broad conformational ensembles we see in bulk bilayers suggest that the structures observed in pure lipid crystals occur seldomly in liquid state. Consequently, in protein binding the entropic cost on the lipid may be significant.
Applications of the proposed model in molecular cell biology include the understanding of lipid-mediated cell signaling
and functions of lipid regulated membrane proteins in general. 
In bionanotechnology, our model facilitates, for example,
the design of phoshopolipid-specific antibodies and lipid nanoparticle carriers with minimal sides effects for mRNA vaccines.
Furthermore, our work has demonstrated the power of open access MD simulation data
to complement the PDB data in elucidating behavior of complex systems of disordered biomolecules.
